---
title: "The impact of government expenditure on education on income inequality in the MENA region: Panel analysis on the time period from 1990 to 2017."
output:
  pdf_document: default
  html_document:
    pdf_print: 
        keep_tex: true
      includes:
          \usepackage{dcolumn} 
  word_document: default
---

# Description:

This paper analyzes the effect of government educational spending has on income inequality .

Dependent variable : Income inequality represented by the GINI index

Independent Variable: Government expenditure on education, total (% of GDP)

Control Variables:

-   Unemployment, total (% of total labor force)

-   GDP per capita growth (annual %)

-   Age dependency ratio (% of working-age population)

-   Population growth (annual %)

-   School enrollment, primary (% gross)

-   School enrollment, secondary (% gross)

-   School enrollment, tertiary (%gross)

-   Urban population (% of total population)

-   Population density (people per sq. km of land area)

-   Agriculture, forestry, and fishing, value added (% of GDP)

-   Industry (including construction), value added (% of GDP)

-   Consumer price index (2010 = 100)

-   Urban population (% of total population)

# Methodology:

Our goal is to fit a significant model that captures the magnitude of the effect government education expenditure has on income inequality, the latter of which is represented by the Gini index. Whereas a value of 0 for the Gini index implies a perfect income distribution and 100% implies perfect income inequality.

Pooled OLS is the simplest way of estimating any panel data set however its assumptions must be met . This study chose Pooled estimation not only for its simplicity but also due the small data set at hand . It could be assumed that the model will not suffer from autocorrelation since observations weren't recorded in the exact same period and they're of limited quantity .In order to achieve a statistically viable model all the variables that affect income inequality has to be taken into consideration. However some of these variables are not quantifiable thus, we selected a group variable that are proven statistically significant. Which are as follows: unemployment rate, population growth, GDP per capita, and educational enrollment for primary, secondary, and tertiary Vazquez et al. (2012). And according to Kaasa (2003), there are about 14 variables that explain 68.6% of variability of the GINI index. Namely; unemployment, inflation, Government expenditure on education and health as a percentage of GDP, school enrolment (primary), growth of population, population density.

## Variable Descriptions

## Imports

```{r echo = T, results = 'hide'}

library(readxl)
library(ggplot2)
library(dplyr)
library(summarytools)
library(corrplot)
library(mice)
library(rstatix)
library(MVN)
library(bestNormalize)
library(car)
library(lmtest)
library(plm)
library(leaps)
library(tseries)
library(dynlm)
library(sjPlot)
library(sjmisc)
library(sjlabelled)
library(xtable)
library(stargazer)
```

```{r}
df <- read_excel("Final_Data1.xlsx") # total observations 532
df_omitted <- na.omit(df) # Non-Na observations 37 
```

### Imputing missing data points

MICE stands for Multivariate Imputation by Chained Equations. this procedure replaces missing values with plausible values in order to not affect the regression coefficients .

```{r echo = T, results = 'hide'}
imputed_Data <- mice(df, m=5, maxit = 50, method = 'pmm', seed = 500)
imputed_Data <- complete(imputed_Data,3)

```

## Descriptive statistics

```{r}
df_desc <- imputed_Data[-c(1,2) ] # Dropped Country and time  column
descr(df_desc)
```

### Correlations

```{r}
df_corr <- cor(df_desc)
corrplot(df_corr, method="circle"  )
```

From the correlogram above , it is observed that the there are correlation between our dependent variable and some explanatory variables, which will lay the framework for the OLS model . Moreover, there appears to be high correlations between the explanatory variables , which might violate the multicolinearity assumption later on the Modelling section.

All variables are not normally distributed except the AGE variable (Age dependency ratio). Unfortunately, Log , square root , Box-Cox ,and reciprocal transformation cant be used as our observations include non-positive values.

#### Encoding

GDP per capita growth and population growth rate both have negative values , which complicates the transformation process . The following procedure will transform both those variables into categorical variables signifying High , Medium ,and Low growth rates , relative to the MENA region.

```{r}
df_desc$GDPC <- cut(df_desc$GDPC,
              breaks=c(-65, -15, 35, 100),
              labels=c('Low', 'Medium ', 'High '))
df_desc$POP <- cut(df_desc$POP,
              breaks=c(-5, 3, 11, 19),
              labels=c('Low', 'Medium ', 'High '))
```

## Modelling and Feature selection

```{r}
#df_omitted1 <- df_omitted[-c(1,2)] #  Radj= 0.92 , xd
lm1 <- lm(GINI ~ .,data=df_desc) 
summary(lm1)
plot(lm1)

```

### Assumptions

1.  **Normality**

2.  **Homoscedasticity**

    $$ Var(E_i) = \sigma^2 $$

3.  **Autocorrelation**

    $$Cov(E_i,E_j)=0$$

4.  **Multicolinearity**

### Normality

```{r}
hist(df_desc$EXP,
main="EXP Distribution ",
xlab="Government expenditure on education, total (% of GDP)",
freq=FALSE
)
```

```{r}
#hist(sqrt(df_desc1$GINI),
#main="Gini index Distribution",
#xlab="Gini index",
#freq=FALSE
#)
```

```{r}
Ntest <- mvn(data = df_desc[-c(4,6)], mvnTest = "mardia")
Ntest$univariateNormality
```

#### Transformation

Unfortunately, no universal transformation will suffice . Thus , each variable has to be transformed separately . The interpretation of a box cox transformed variable is simply X\^ lambda. I

It is important to note that while the tests show that the normality assumption has been violated , histograms of the variables appear to follow a normal distribution.

```{r}
df_desc1 <- df_desc[-c(4,6)]
df_matrix <- as.matrix(df_desc[-c(4,6)])

df_desc1[1] <- sqrt(df_desc$GINI) #GINI normalize

df_desc1[2] <-log10(df_desc1[2]+0.05) #GOV expenditure normalized , log base 10 normalizes positively skewed data
df_desc1[3] <- sqrt(df_desc1[3]) #unemployment normalized
bc<-boxcox(imputed_Data$PRIM)
  df_desc1[5] <-bc$x.t #Primary enrollment normalized,  lambda = 1.999958
bc_sec <- boxcox(df_desc1$SEC)
df_desc1[6] <-bc_sec$x.t #Secondary enrollment normalized,  lambda =  1.756894
df_desc1[7] <- sqrt(df_desc1[7]) #tertiary enrollment normalized
bc_popd <- boxcox(df_desc1$PopD)
df_desc1[9] <- bc_popd$x.t #Population density  normalized,  lambda =  0.02170782
bc_GDPAFF <- boxcox(df_desc1$GDPAFF)
df_desc1[10] <- bc_GDPAFF$x.t #Population density  normalized,  lambda =  0.3059507
df_desc1[11] <-log(df_desc1[11]) #GDP Industry expenditure normalized 
df_desc1[12] <- sqrt(df_desc1[12]) #CPI normalized
bc_HealthGPD <- boxcox(df_desc1$HealthGPD)
df_desc1[13] <- bc_HealthGPD$x.t #Health GDP expenditures density  normalized, lambda = 0.1132915
df_desc1 <- cbind(df_desc1,df_desc[c(4,6)] )

```

### Homoscedasticity

Breusch-Pagan Test is used to determine if Homoscedasticity is present in the analysis.

$$
H_1 : Var(E_i) 
\neq
\sigma^2$$

$$
H_1 : Var(E_i) 
\neq
\sigma^2$$

```{r}
bptest (lm1)
```

P-value is more than specified alpha (0.05) , thus there is no enough evidence to reject the null hypothesis Homoscedasticity is present

### Autocorrelation

Durbin-Watson test is used to determine if the errors are autocorrelated.

$$
H_0 : Cov(E_i,E_j)=0
$$

$$
H_1 : Cov(E_i,E_j)
\neq 0 
$$

```{r}
durbinWatsonTest(lm1)
```

There is enough evidence to reject the null hypothesis and conclude that the errors are correlated .

The literature suggests that government effectiveness and corruption has a significant effect on income inequality . However , no estimates for either are available consistently throughout the MENA region . We can conclude that the autocorrelation present is caused by the missing variable , commonly referred to as omitted variable bias .

### Multicolinearity

Variance inflation factor or VIF for short is a statistic computed to determine which regressor is the main cause of multicolinearity in our model . A relatively high VIF is a good sigh to drop that variable .

```{r}
vif(lm1)
```

GDPAFF variable has the greatest VIF out of all the variables which is reaffirmed by the correlogram conducted above .

### Re-fitting

```{r}
df_fixed <-df_desc1[-c(10)]
lm2 <- lm(GINI ~ . ,data=df_fixed) 
summary(lm2)
plot(lm2)
```

### Feature Selection

```{r}
subs <- regsubsets(GINI ~ . ,data=df_fixed, nvmax = 15)

res.sum <- summary(subs)
table(res.sum$bic)
```

According to the BIC , Mallow's C , and adjusted R-squared the following model has the optimal combination of variables :

```{r}
lm3 <- lm(GINI ~ . ,data=df_fixed[-c(5)]) #drop prim
summary(lm3)
```

## Robustness

```{r}
lm4 <- lm(GINI ~ EXP   +SEC  + TERT  +PopD  + CPI    , data= df_fixed) # Without unemployment 
summary(lm4)
```

```{r}
lm5 <- lm(GINI ~ EXP  + UNP  + AGE + TERT + URB +PopD +GDPI + CPI +HealthGPD +GDPC + POP , data= df_fixed) # Without Secondary edu 
summary(lm5)
```

```{r}
lm6 <- lm(GINI ~ EXP  + UNP +SEC + TERT + URB +PopD +GDPI + CPI +HealthGPD +GDPC + POP , data= df_fixed) # Without AGE
summary(lm6) 
```

```{r}
lm7 <- lm(GINI ~ EXP  + UNP +SEC + AGE  + URB +PopD +GDPI + CPI +HealthGPD +GDPC + POP , data= df_fixed) # Without TERTiary
summary(lm7)
```

```{r}
df_fixed$LGINI <- Lag(df_fixed$GINI,-1)
lm8 <- lm(GINI ~ EXP  + UNP +SEC + AGE  + TERT +PopD +GDPI + CPI +HealthGPD +GDPC + POP + URB +LGINI , data= df_fixed) # Without URbAn
summary(lm8)
```

```{r}
lm9 <- lm(GINI ~ EXP  + UNP +SEC + AGE  + TERT +URB +GDPI + CPI +HealthGPD +GDPC + POP , data= df_fixed) # Without PopD
summary(lm9)
```

```{r}
lm10 <- lm(GINI ~ EXP  + UNP +SEC + AGE  + TERT +URB +PopD + CPI +HealthGPD +GDPC + POP , data= df_fixed) # Without GDPI
summary(lm10)
```

```{r}
lm11 <- lm(GINI ~ EXP  + UNP +SEC + AGE  + TERT +URB +PopD + GDPI +HealthGPD +GDPC + POP , data= df_fixed) # Without CPI
summary(lm11)
```

```{r}
lm12 <- lm(GINI ~ EXP  + UNP +SEC + AGE  + TERT +URB +PopD + GDPI +CPI +GDPC + POP , data= df_fixed) # Without HealthGPD   
summary(lm12)
```

```{r}
lm13 <- lm(GINI ~ EXP  + UNP +SEC + AGE  + TERT +URB +PopD + GDPI +CPI +HealthGPD + POP , data= df_fixed) # Without GDPC   
summary(lm13)
```

```{r}
lm14 <- lm(GINI ~ EXP  + UNP +SEC + AGE  + TERT +URB +PopD + GDPI +CPI +HealthGPD + GDPC , data= df_fixed) # Without POP   
summary(lm14)
```

```{r}
tab_model(lm3, lm4, lm5,lm6, lm7, lm8,lm9, lm10, lm11,lm12, lm13, lm14 , auto.label = FALSE)
```

```{r}
tab_model(
  lm3, lm14, lm13, 
  pred.labels = c("Intercept"
                  , "Government expenditure on education, total (% of GDP)"
                  , "Unemployment, total (% of total labor force)"
                  , "Age dependency ratio (% of working-age population)"
                  ,"School enrollment, secondary (% gross)"
                  , "School enrollment, tertiary (% gross)"
                  , "Urban population (% of total population)" 
                  ,"Population density (people per sq. km of land area)"
                  ,"Industry (including construction), value added (% of GDP)"
                  ,"Consumer price index (2010 = 100)"
                  , "HealthGPD" ,"GDP per capita(Moderate)"
                  ,"GDP per capita(High)" 
                  , "Population Growth rate(Moderate)" 
                  , "Population Growth rate (high)" ),
  dv.labels = c("First Model", "Controlling for HealthGPD", "Controlling for Population Growth rate"),
  string.pred = "Coeffcient",
  string.ci = "Conf. Int (95%)",
  string.p = "P-Value"
)
```

```{r}
lm15 <- lm(GINI ~ EXP +SEC +POP +GDPI +URB +HealthGPD +TERT +PopD , data= df_fixed)
lm16 <- lm(GINI ~ EXP +SEC  +GDPI +URB +HealthGPD +TERT +PopD , data= df_fixed)
lm17 <- lm(GINI ~ EXP +SEC +POP  +URB +HealthGPD +TERT +PopD , data= df_fixed)
lm18 <- lm(GINI ~ EXP +SEC +POP +GDPI  +HealthGPD +TERT +PopD , data= df_fixed)
lm19 <- lm(GINI ~ EXP +SEC +POP +GDPI +URB  +TERT +PopD , data= df_fixed)
lm20 <- lm(GINI ~ EXP +SEC +POP +GDPI +URB +HealthGPD  +PopD , data= df_fixed)
lm21 <- lm(GINI ~ EXP +SEC +POP +GDPI +URB +HealthGPD +TERT  , data= df_fixed)

summary(lm15)
```

```{r}
stargazer(lm15, lm21, lm16,lm17,lm18,lm19,lm20, title="Results", align=TRUE, type = "latex")
```

```{=latex}
% Table created by stargazer v.5.2.3 by Marek Hlavac, Social Policy Institute. E-mail: marek.hlavac at gmail.com
% Date and time: Sun, May 07, 2023 - 8:57:57 PM
% Requires LaTeX packages: dcolumn 
\begin{table}[!htbp] \centering 
  \caption{Results} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lD{.}{.}{-3} D{.}{.}{-3} D{.}{.}{-3} D{.}{.}{-3} D{.}{.}{-3} D{.}{.}{-3} D{.}{.}{-3} } 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{7}{c}{\textit{Dependent variable:}} \\ 
\cline{2-8} 
\\[-1.8ex] & \multicolumn{7}{c}{GINI} \\ 
\\[-1.8ex] & \multicolumn{1}{c}{(1)} & \multicolumn{1}{c}{(2)} & \multicolumn{1}{c}{(3)} & \multicolumn{1}{c}{(4)} & \multicolumn{1}{c}{(5)} & \multicolumn{1}{c}{(6)} & \multicolumn{1}{c}{(7)}\\ 
\hline \\[-1.8ex] 
 EXP & -0.279 & -0.446^{*} & -0.332^{} & -1.022^{*} & -0.443^{*} & -0.467^{} & -0.450^{**} \\ 
  & (0.177) & (0.177) & (0.176) & (0.183) & (0.200) & (0.182) & (0.173) \\ 
  & & & & & & & \\ 
 SEC &  & -0.414^{*} & -0.404^{*} & -0.527^{*} & -0.187^{*} & -0.466^{*} & -0.288^{*} \\ 
  &  & (0.055) & (0.055) & (0.058) & (0.060) & (0.056) & (0.040) \\ 
  & & & & & & & \\ 
 POPMedium  & -0.531^{*} & -0.597^{*} &  & -0.645^{*} & -0.376^{*} & -0.640^{*} & -0.561^{*} \\ 
  & (0.083) & (0.081) &  & (0.087) & (0.092) & (0.083) & (0.080) \\ 
  & & & & & & & \\ 
 POPHigh  & -0.929^{*} & -0.822^{*} &  & -0.965^{*} & -0.480 & -0.974^{*} & -0.879^{**} \\ 
  & (0.335) & (0.331) &  & (0.354) & (0.374) & (0.339) & (0.321) \\ 
  & & & & & & & \\ 
 GDPI & -1.090^{*} & -0.770^{*} & -1.034^{*} &  & -0.799^{*} & -1.008^{*} & -1.037^{*} \\ 
  & (0.086) & (0.079) & (0.086) &  & (0.097) & (0.089) & (0.080) \\ 
  & & & & & & & \\ 
 URB & 0.040^{*} & 0.043^{*} & 0.044^{*} & 0.042^{*} &  & 0.033^{*} & 0.044^{*} \\ 
  & (0.003) & (0.003) & (0.003) & (0.003) &  & (0.003) & (0.003) \\ 
  & & & & & & & \\ 
 HealthGPD & -0.449^{*} & -0.384^{*} & -0.451^{*} & -0.434^{*} & -0.044 &  & -0.366^{*} \\ 
  & (0.048) & (0.048) & (0.048) & (0.051) & (0.047) &  & (0.046) \\ 
  & & & & & & & \\ 
 TERT & -0.062^{*} & 0.083^{*} & 0.121^{*} & 0.196^{*} & -0.003 & 0.034 &  \\ 
  & (0.022) & (0.030) & (0.030) & (0.031) & (0.033) & (0.030) &  \\ 
  & & & & & & & \\ 
 PopD & -0.213^{*} &  & -0.234^{*} & -0.038 & -0.100^{*} & -0.183^{*} & -0.194^{**} \\ 
  & (0.036) &  & (0.036) & (0.035) & (0.040) & (0.037) & (0.035) \\ 
  & & & & & & & \\ 
 Constant & 1.550^{*} & -0.372 & 0.117 & -3.193^{*} & 3.252^{*} & 1.485^{*} & 0.870^{*} \\ 
  & (0.386) & (0.425) & (0.436) & (0.339) & (0.420) & (0.408) & (0.320) \\ 
  & & & & & & & \\ 
\hline \\[-1.8ex] 
Observations & \multicolumn{1}{c}{644} & \multicolumn{1}{c}{644} & \multicolumn{1}{c}{644} & \multicolumn{1}{c}{644} & \multicolumn{1}{c}{644} & \multicolumn{1}{c}{644} & \multicolumn{1}{c}{644} \\ 
R$^{2}$ & \multicolumn{1}{c}{0.395} & \multicolumn{1}{c}{0.415} & \multicolumn{1}{c}{0.404} & \multicolumn{1}{c}{0.329} & \multicolumn{1}{c}{0.252} & \multicolumn{1}{c}{0.380} & \multicolumn{1}{c}{0.435} \\ 
Adjusted R$^{2}$ & \multicolumn{1}{c}{0.388} & \multicolumn{1}{c}{0.408} & \multicolumn{1}{c}{0.397} & \multicolumn{1}{c}{0.321} & \multicolumn{1}{c}{0.243} & \multicolumn{1}{c}{0.373} & \multicolumn{1}{c}{0.428} \\ 
Residual Std. Error & \multicolumn{1}{c}{0.783 (df = 635)} & \multicolumn{1}{c}{0.770 (df = 635)} & \multicolumn{1}{c}{0.776 (df = 636)} & \multicolumn{1}{c}{0.824 (df = 635)} & \multicolumn{1}{c}{0.870 (df = 635)} & \multicolumn{1}{c}{0.792 (df = 635)} & \multicolumn{1}{c}{0.756 (df = 635)} \\ 
F Statistic & \multicolumn{1}{c}{51.851$^{*}$ (df = 8; 635)} & \multicolumn{1}{c}{56.288$^{*}$ (df = 8; 635)} & \multicolumn{1}{c}{61.586$^{*}$ (df = 7; 636)} & \multicolumn{1}{c}{38.942$^{*}$ (df = 8; 635)} & \multicolumn{1}{c}{26.757$^{*}$ (df = 8; 635)} & \multicolumn{1}{c}{48.715$^{*}$ (df = 8; 635)} & \multicolumn{1}{c}{61.106$^{*}$ (df = 8; 635)} \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{7}{r}{$^{}$p$<$0.1; $^{}$p$<$0.05; $^{**}$p$<$0.01} \\ 
\end{tabular} 
\end{table}
```
